
This chapter is based on the following article:
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%
\item \textbf{Dobbelaere, T.}, Curcic, M., Le Hénaff, M., \& Hanert, E. (2022). Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes. \textit{Ocean Modelling}, 101947. \href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.1016/j.ocemod.2022.101947}.
\end{list}

\begin{abstract}
    The intensity of major tropical cyclones has increased during the past decade. Their effect is particularly acute in coastal areas where they cause extensive damage leading to an influx of debris, sediments and waste to the sea. However, most operational coastal ocean models do not represent heavy-wind transport processes correctly \modif{if} the hydrodynamics is not coupled with the wind-generated waves. This may lead to significant errors in ocean simulations under tropical cyclone conditions. Here, we investigate current-wave interactions during a major hurricane and assess their impact on transport processes. We do that by coupling the unstructured-mesh coastal ocean model SLIM with the spectral wave model SWAN, and applying it to the Florida Reef Tract during Hurricane Irma (September 2017). We show that the coupled model successfully reproduces the wave behavior, the storm surge and the ocean currents during the passage of the hurricane. We then use the coupled and uncoupled wave-current model to simulate the transport of passive drifters. We show that the wave radiation stress gradient alone can lead to changes of up to 1 m/s in the modeled currents, which in turn leads to differences of up to 5 km in the position of drifting material over the duration of the hurricane. The Stokes drift however appears to cause deflections up to 4 times larger and hence dominates wave-induced transport. Wave-current interactions therefore strongly impact the transport of drifting material such as sediments and debris in the aftermath of a hurricane. They should thus be taken into account in order to correctly assess its overall impact.
\end{abstract}

%\newpage



\input{chapters/irma/article}